Expanding the scope of experimental archaeology using the Perception-Process-Product conceptual framework

### 2023-03-14 manuscript/manuscript.Rmd
minor revision.

This paper presents the Perception-Process-Product (hereinafter referred to as "Triple P") conceptual framework to expand the scope of experimental archaeology, which tends to centers around the reverse engineering of a past technology in a minimal or least-effort manner while ignoring the rich contextual information it affords. The Triple P framework aims to amplify the expression of variability in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) and better identify the complex interacting relationships across these three levels of variations (**Figure** \@ref(fig:concept)). To accomplish these two objectives, we advocate the following four measures: 1) normalizing the ethological and ethnographic data collection in experimental projects ; 2) encouraging the involvements of avocational as well as novice participants; 3) boosting the collaboration across labs in a global scale; 4) building an open-access repository for data reuse. It is no doubt that strategies of data collection and analysis of a given experimental project should be primarily derived from the research question, which can be legitimately narrow in scope, but the awareness of the rich toolkit available can sometimes inspires researchers to ask questions that are bold and transformative [@schmidt2020]. Here I will mainly leverage the extensive corpus in experimental designs and inferences revolving around stone artifacts to demonstrate the necessity and potentials of this framework.

@@

One of the major concerns of experimental archaeology design, as in all empirical social sciences, is the validity, namely how good is a particular conclusion or inference approximates the true condition. The concept of validity has multiple dimensions, and one of the most commonly used classification schemes is internal versus external validity. Roe and Just (2009: 1266-1267) defined internal validity as “the ability of a researcher to argue that observed correlations are causal” and external validity as “the ability to generalize the relationships found in a study to other persons, times, and settings”. This balance between these two validity concepts is an issue that cannot be escaped for all experimental archaeology project designs. In the context of stone tool replication, it can be projected into the debate over the use of machines in knapping, a research design that has received increasing attention in the past decades (Eren et al., 2016). Machine knapping is a typical design with high internal validity but low external validity, which has been proved to provides critical insights into lithic fracture mechanics by identifying potential causal variables at the level of individual stone artifact such as determining which angle of blow or how much force of blow will produce the maximal amount of blade area. All the variables of interest are easy to measure, quantify, and control in a machine knapping setting. Nevertheless, being easy to control is not always a virtue as it essentially eliminates the potential interactions between variables operable in the past and thereby providing misleading results when answering archaeological questions. In addition to the applications of machine knapping, the same problem is also incurred by the introduction of standardized artificial material like bricks or video instruction in teaching experiments. As a rule of thumb, external validity should be given more weight in the design when the research focuses on the behaviors of the users of artifacts while internal validity matters more when it comes to the properties of artifacts themselves.

@@

```{r concept, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="The conceptual diagram of the Perception-Process-Product analytical framework."}
knitr::include_graphics(here::here("figure/Fig1.jpg"))
```

>>>>

This paper presents the Perception-Process-Product (hereinafter referred to as "Triple P") conceptual framework to expand the scope of experimental archaeology, which tends to center around the reverse engineering of a past technology in a minimal or least-effort manner while ignoring the rich contextual information it affords. The Triple P framework aims to **a)** amplify the expression of variability in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) and **b)** better identify the complex interacting relationships across these three levels of variations (**Figure** \@ref(fig:concept)). To accomplish these two objectives, we advocate the following four measures: 1) normalizing the ethological and ethnographic data collection in experimental projects ; 2) encouraging the involvements of avocational as well as novice participants; 3) boosting the collaboration across labs in a global scale; 4) building an open-access repository for data reuse. It is no doubt that strategies of data collection and analysis of a given experimental project should be primarily derived from the research question, which can be legitimately narrow in scope, but the awareness of the rich toolkit available can sometimes inspires researchers to ask questions that are bold and transformative [@schmidt2020]. Here I will mainly leverage the extensive corpus in experimental designs and inferences revolving around stone artifacts to demonstrate the necessity and potentials of this framework.

@@

One of the major concerns of experimental archaeology design, as in all empirical social sciences, is the validity, namely how good is a particular conclusion or inference approximates the true condition. The concept of validity has multiple dimensions, and one of the most commonly used classification schemes is internal versus external validity. Roe and Just (2009: 1266-1267) defined internal validity as "the ability of a researcher to argue that observed correlations are causal" and external validity as "the ability to generalize the relationships found in a study to other persons, times, and settings". This balance between these two validity concepts is an issue that cannot be escaped for all experimental archaeology project designs. In the context of stone tool replication, it can be projected into the debate over the use of machines in knapping, a research design that has received increasing attention in the past decades (Eren et al., 2016). Machine knapping is a typical design with high internal validity but low external validity, which has been proved to provides critical insights into lithic fracture mechanics by identifying potential causal variables at the level of individual stone artifact such as determining which angle of blow or how much force of blow will produce the maximal amount of blade area. All the variables of interest are easy to measure, quantify, and control in a machine knapping setting. Nevertheless, being easy to control is not always a virtue as it essentially eliminates the potential interactions between variables operable in the past and thereby providing misleading results when answering archaeological questions. In addition to the applications of machine knapping, the same problem is also incurred by the introduction of standardized artificial material like bricks or video instruction in teaching experiments. As a rule of thumb, external validity should be given more weight in the design when the research focuses on the behaviors of the users of artifacts while internal validity matters more when it comes to the properties of artifacts themselves.

@@

```{r concept, echo=FALSE, out.width = '80%', fig.align='center', fig.cap="A schematic diagram of the Perception-Process-Product conceptual framework."}
knitr::include_graphics(here::here("figure/Fig1.jpg"))
```
